FantasticLamp: a genome graph based pipeline for calculating the efficacy of genomic edits

### 2023-08-03 paper/main.tex
ok but where are the experiments

\abstract{
\textbf{Motivation:}
Accurately calculating the efficacy of genomic edits is crucial to understanding the performance of the editing techniques, in order to optimize methods and improve the success rate of the edits.
Additionally, by understanding the success rate of genomic edits, researchers can identify any potential problems or limitations of the techniques and work to overcome them.
This can help to improve the accuracy and precision of the editing methods, which is essential for many applications, such as creating genetically engineered cells for therapeutic purposes, understanding gene functions, and studying genetic diversity in a population of cells. Current linear alignment methods for quantifying the success of edits suffer from reference bias and fail to capture the complexity involved when dealing with multiple edit states.\\
% Mention current methods and why there is need for better methods.
\textbf{Results:}
We design \textit{FantasticLamp}, an open source genome-graph-based pipeline for calculating the efficacy of genomic edits performed on multiple populations of cells.
It constructs a variation graph from the reference genome and edit template sequences that includes both edited and unedited sequences as paths, and combinations of edits as potential walks through the graph.
It then maps reads from the edited populations onto this graph in order to calculate the coverage of the edited sequences compared to the unedited sequences.
This pipeline aims to provide a quantitative measure of the success of each genomic edit without bias towards the reference or particular single editing constructs.\\
% Reword this paragraph, add detail regarding the process. Mention the successful implementation.
\textbf{Availability:}
\textit{FantasticLamp} is published as free software under the MIT open source license.
Source code and documentation are available at \url{https://github.com/casper-schutte/fantastic-lamp}.
\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
}

@@

The pipeline then maps reads from the edited populations to this graph using \textit{vg} \citep{garrison2018variation}, generating a GAF (Gene Annotation Format) file to represent the alignment.
A Python script parses the alignment file to calculate the coverage of the homology arms relative to the reference homology arms.
We output a coverage table (in TSV), displaying each intended edit sequence alongside the corresponding homology arm coverage and reference homology arm coverage, thus quantifying the efficacy of each edit.

@@

Preliminary results demonstrated the pipeline's capacity to calculate the read coverage, which acted as a measure of the effectiveness of the implemented edits.
However, due to unforeseen logistical constraints, further usage of this dataset was halted.
Consequently, an artificial dataset was constructed for further testing and refinement.

@@

\section*{Discussion}
\label{sec:discussion}
\textit{FantasticLamp} demonstrates using a variation graph to quantify complex genome edits.
We consider this work as an example of how to apply unbiased edit quantification to editing experiments.
However, much work is left to be done to use these approaches at scale.
As an emerging field, commodity high-throughput genome editing is still rare and we lack accepted standards for experimental design, so our implementation's inputs and outputs are necessarily bespoke.
While we demonstrate basic functionality of \textit{FantasticLamp} on artificial and singleplex data, it could be expanded to analyze multiplexed pooled data.
This pipeline provides a starting point to avoid bias when quantifying edits that can be built upon as editing technologies advance.
Potential improvements include handling more complex structural variants, incorporating multiplexed data, estimating off-target effects, optimizing runtime for large genomes, and interfacing for non-expert users.
In summary, \textit{FantasticLamp} provides a basic demonstration of the principle of using a variation graph as a reference system to avoid bias when quantifying genomic edits.

>>>>

\abstract{
\textbf{Motivation:}
Accurately calculating the efficacy of genomic edits is crucial to understanding the performance of the editing techniques, in order to optimize methods and improve the success rate of the edits.
Additionally, by understanding the success rate of genomic edits, researchers can identify any potential problems or limitations of the techniques and work to overcome them.
This can help to improve the accuracy and precision of the editing methods, which is essential for many applications, such as creating genetically engineered cells for therapeutic purposes, understanding gene functions, and studying genetic diversity in a population of cells.
Current linear alignment methods for quantifying the success of edits exhibit reference bias and cannot model complex mixed edit states.\\
% Mention current methods and why there is need for better methods.
\textbf{Results:}
We design \textit{FantasticLamp}, an open source genome-graph-based pipeline for calculating the efficacy of genomic edits performed on multiple populations of cells.
It constructs a variation graph from the reference genome and edit template sequences that includes both edited and unedited sequences as paths, and combinations of edits as potential walks through the graph.
It then maps reads from the edited populations onto this graph in order to calculate the coverage of the edited sequences compared to the unedited sequences.
This pipeline aims to provide a quantitative measure of the success of each genomic edit without bias towards the reference or particular single editing constructs.\\
% Reword this paragraph, add detail regarding the process. Mention the successful implementation.
\textbf{Availability:}
\textit{FantasticLamp} is published as free software under the MIT open source license.
Source code and documentation are available at \url{https://github.com/casper-schutte/fantastic-lamp}.
\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
}

@@

The pipeline then maps reads from the edited populations to this graph using \textit{vg} \citep{garrison2018variation}, generating a GAF (Gene Annotation Format) file to represent the alignment.
A Python script parses the alignment file to calculate the coverage of the homology arms relative to the reference homology arms.
We output a coverage table (in TSV format), displaying each intended edit sequence alongside the corresponding homology arm coverage and reference homology arm coverage, thereby quantifying the efficacy of each edit.

@@

We found that the pipeline captured expected edit states.
However, due to unforeseen logistical constraints, further usage of this dataset was halted.
Consequently, an artificial dataset was constructed for further testing and refinement.

@@

\section*{Discussion}
\label{sec:discussion}
\textit{FantasticLamp} demonstrates using a variation graph to quantify complex genome edits.
We consider this work as an example of how to apply unbiased edit quantification to editing experiments.
However, much work is left to be done to use these approaches at scale.
%As an emerging field, commodity high-throughput genome editing is still rare and we lack accepted standards for experimental design, so our implementation's inputs and outputs are necessarily bespoke.
As an emerging field, commodity high-throughput genome editing is still uncommon, and accepted standards for experimental design are lacking.
Therefore, the inputs and outputs of our implementation are necessarily customized.

This pipeline provides a starting point to avoid bias when quantifying edits that can be built upon as editing technologies advance.
While we demonstrate basic functionality of \textit{FantasticLamp} on artificial and singleplex data, it could be expanded to analyze multiplexed pooled data.
Other potential improvements include handling more complex structural variants, incorporating multiplexed data, estimating off-target effects, and optimizing runtime for large genomes.
In summary, \textit{FantasticLamp} provides a basic demonstration of the principle of using a variation graph as a reference system to avoid bias when quantifying genomic edits.
